The \dealii{} Library, Version 9.3

### 2021-04-23 9.3/paper.tex
Describe step-75
Write a bit more material.

\subsubsection{Miscellaneous}

@@

Until now, \dealii has only supported local smoothing multigrid algorithms~\citep{ClevengerHeisterKanschatKronbichler2019}. With release 9.3, the support for global coarsening of continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP}) elements  in a geometric and a polynomial sense has been added. These multigrid variants promise less iteration numbers and better parallel scalability with the disadvantage of having to deal with hanging nodes during smoothing and of more work per iteration overall.

Global-coarsening algorithms~\citep{becker00} either coarsen each cell of a triangulation or reduce the polynomial degree of the finite element of each cell and hereby obtain a new system with less unknowns (also known as "coarse level").
Repeating such kind of coarsening results in a sequence of multigrid levels.

The transfer between two levels has been encoded in the new class \texttt{MGTwoLevelTransfer}, which can be set up via the functions \texttt{MGTwoLevelTransfer::\allowbreak reinit\_\allowbreak geometric\_\allowbreak transfer()} or \texttt{MGTwo\allowbreak LevelTransfer::\allowbreak reinit\_\allowbreak polynomial\_\allowbreak transfer()} for given
\texttt{DoFHandler} and \texttt{AffineConstraint} classes of two levels. The resulting transfer operators
between two levels can be collected in a single object
(\texttt{MGTransferGlobalCoarsening}) that can be used just as \texttt{MGTransferMatrixFree} and can be passed to the actual \texttt{Multigrid}
algorithm. The option that users can set up the transfer operators between two levels on their own enables
the mix of polynomial  and geometric coarsening, as needed. It should also be noted that a transfer between continuous
and discontinuous elements is possible by providing \texttt{DoFHandler}s set up with this kind of elements.

Internally, the transfer operators categorize fine/coarse cell pairs according to the
occurrence of refinement
(refinement or no-refinement) and to the polynomial-degree pair ($k_f$-$k_c$) into groups.
The projection and restriction are performed with efficient matrix-free kernels applying
1D projection and restriction matrices in the case of hypercube cells and via full projection and
restriction matrices in the case of simplex cells.

The user is responsible for setting up the levels (including the operators and the
smoothers). Please note that global coarsening can currently only be performed between active levels.
The user can apply utility
functions from the \texttt{MGTransferGlobalCoarseningTools} namespace to set up the levels. For
constructing the diagonal or the system matrix for a matrix-free operator,
needed for the smoother or the coarse-grid solver,
the functions \texttt{create\_diagonal()} and \texttt{create\_matrix()} from
the \texttt{MatrixFreeTools} namespace can be used (see also Subsection~\ref{subsec:mf}).

The usage of the new transfer operators (and of some of the utility functions) in the context of a hybrid multigrid algorithm ($hp$-MG with AMG as coarse-grid solver) for $hp$-problems is demonstrated in the new tutorial \texttt{step-75} (see also Subsection~\ref{subsec:steps}).

@@

The classes \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} are highly
templated to reach high performance. In particular, the template parameters include
the polynomial degree of the finite element $k$ and the number of the 1D quadrature points $q$.
For application codes that rely on operators of many different degrees (e.g., because
they use $p$-multigrid or $hp$-algorithms), the instantiation
means a long compilation time.

In the current release, we have improved the \texttt{FEEvaluation} and the \texttt{FEFaceEvaluation} class implementations that do
not rely on the template parameters $k$ and $q$ (expressed in the code
with "-1" and "0"):
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
The meaning of the new first argument is
discussed in Subsubsection~\ref{subsubsection:mf:hp}.

These classes select at runtime---for common low and medium polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1, k+2, \lfloor (3k)/2 \rfloor \}$)---efficient precompiled implementations and default to non-templated
evaluation kernels else.

@@

With release 9.1, large parts of the $hp$-algorithms in \dealii were parallelized so that
parallel matrix-based simulations can be performed with the $hp$-infrastructure. In the present
release, we have extended the setup routines of \texttt{MatrixFree} so that it now also
provides parallel $hp$-support.

@@

The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes are extended for the $hp$-case. To nevertheless determine
the active FE and quadrature index, the current cell/face range has to be provided
the constructors of the \texttt{FEEvaluation} classes, from which the relevant information
can be deduced (in the simplex case also the face type). These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

\subsubsection{MPI-3.0 shared-memory support}
The classes \texttt{MatrixFree} and \texttt{LinearAlgebra::\allowbreak distributed::\allowbreak Vector} have been extended with
MPI-3.0 shared-memory features. If \texttt{MatrixFree} has been configured by setting
\texttt{MatrixFree::AdditionalData::communicator\_sm} by an appropriate user-provided subcommunicator,
vectors are created by \texttt{MatrixFree::create\_dof\_vector()} in such a way that the
memory of processes in this subcommunicator can be accessed. In particular, this means
that the \texttt{FEEvaluation} classes can access the degrees of freedom owned by
other processes and in certain cases local
communication can be skipped. To prevent race conditions, \texttt{MatrixFree} uses local
barriers at the beginning and the end of loops (\texttt{loop()}, \texttt{cell\_loop()}, \texttt{loop\_cell\_centric()}).

The novel tutorial \texttt{step-76} shows the usage of the new MPI-3.0 feature
 in the context of the solution
of the Euler equations. Here, a speed-up of 27\% could be reached compared to the
original version, \texttt{step-67}, by using the new feature.
For more details and a successful application in the library \texttt{hyper.deal}, see \citep{munch2020hyperdeal}.

\subsubsection{Miscellaneous}

The namespace \texttt{MatrixFreeTools} has been introduced. It collects various utility
functions useful in the context of application of \texttt{MatrixFree}. In particular,
it contains different functions to extract quantities related to a matrix, e.g., the
computation of the diagonal or a matrix with a matrix-free approach.

With version 9.2, we have introduced cell-centric loops. They have been extended to unstructured
meshes. The tutorial \texttt{step-67b} shows their usage in the context of the solution
of the Euler equations.

@@

In addition, there are a number of new tutorial programs:
\begin{itemize}
\item \texttt{step-19} TODO

@@

@@

\item \texttt{step-77} TODO

>>>>

\subsubsection{Miscellanea}

@@

Until now, \dealii has only supported local smoothing multigrid
algorithms~\citep{ClevengerHeisterKanschatKronbichler2019}; in local
smoothing algorithms, one only applies the smoother to that part of an
adaptively refined mesh that has cells on a given refinement level,
but not to parts of the mesh that are coarser than that level.
The current release now also has support for global coarsening~\citep{becker00} when
using continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and
discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP})
elements. Global coarsening builds multigrid levels for the entire
domain, coarsening \textit{all} cells of a triangulation regardless of
how many times they have been refined. In addition, the framework now available in
\dealii{} in not only applicable to geometric coarsening, but can also
perform coarsening by reducing the polynomial degree used, for example
to support $hp$-adaptive meshes where it has been shown that one can
efficiently precondition high-order spaces with lower-order finite
element operators.
Finally, the implementation also supports transfer between continuous
and discontinuous elements by providing \texttt{DoFHandler}s set up
with these kinds of elements; the scheme therefore allows
preconditioning of discontinuous schemes with a continuous
discretization with fewer unknowns, for example.

These new multigrid variants promise fewer solver
iterations and better parallel scalability than the existing local
smoothing algorithms, but have to deal with
hanging nodes within each level and general require more computational
work per iteration overall.

The transfer operators between two levels has been implemented in the new class \texttt{MGTwoLevelTransfer}, which can be set up via the functions \texttt{MGTwoLevelTransfer::\allowbreak reinit\_\allowbreak geometric\_\allowbreak transfer()} or \texttt{MGTwo\allowbreak LevelTransfer::\allowbreak reinit\_\allowbreak polynomial\_\allowbreak transfer()} for given
\texttt{DoFHandler} and \texttt{AffineConstraint} classes of two levels. The resulting transfer operators
can then be collected in a single
\texttt{MGTransferGlobalCoarsening} object that can be used just as \texttt{MGTransferMatrixFree} and can be passed to the actual \texttt{Multigrid}
algorithm.
Several common operations are encoded in utility
functions in the \texttt{MGTransferGlobalCoarseningTools}
namespace. In practice, matrix-free methods typically require
constructing the diagonal of the system matrix for smoothing
operations, and the full matrix on the coarse level for the coarse-grid solver;
the functions \texttt{create\_diagonal()} and \texttt{create\_matrix()} from
the \texttt{MatrixFreeTools} namespace can be used to this end (see also Subsection~\ref{subsec:mf}).

The usage of the new transfer operators (and of some of the utility
functions) in the context of a hybrid multigrid algorithm
($hp$-multigrid with algebraic multigrid as coarse-grid solver) for $hp$-problems is demonstrated in the new tutorial \texttt{step-75}, see also Section~\ref{subsec:steps}.

@@

The classes \texttt{FEEvaluation} and \texttt{FEFaceEvaluation} are highly
templated to reach high performance. In particular, the template parameters include
the polynomial degree of the finite element $k$ and the number of the 1D quadrature points $q$.
For application codes that rely on operators of many different degrees (e.g., because
they use $p$-multigrid or $hp$-algorithms), the many instantiations
necessary and their complexity
imply long compile times.

In the current release, we have improved specializations of these classes that do
not rely on the template parameters $k$ and $q$ (expressed in the code
with ``-1'' and ``0'') -- for example:
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
These classes select at runtime -- for common low and medium
polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1,
k+2, \left\lfloor (3k)/2 \right\rfloor \}$) -- efficient precompiled implementations and default to non-templated
evaluation kernels else.

@@

With release 9.1 \cite{dealII91}, large parts of the $hp$-algorithms in \dealii were parallelized so that
parallel matrix-based simulations can be performed with the $hp$-infrastructure. In the present
release, we have extended the setup routines of \texttt{MatrixFree} so that they now also
provide parallel $hp$-support.

@@

The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes are extended for the $hp$-case. To nevertheless determine
the active FE and quadrature index, the current cell/face range has to be provided
to the constructors of the \texttt{FEEvaluation} classes, from which the relevant information
can be deduced (in the simplex case also the face type). These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

\subsection{MPI-3.0 shared-memory support}
In many large computations, certain pieces of data are computed once
and then threated as read-only. If this information is needed by more
than one MPI process, it is more efficient to store this information only
once in shared memory among all processes located on a multicore
node. MPI supports the creation of such shared memory windows since
version 3.0, and deal.II can now use this in the
\texttt{AlignedVector} and \texttt{Table} classes that are often used
for large lookup tables.

The feature is also used in the \texttt{MatrixFree} and
\texttt{LinearAlgebra::\allowbreak distributed::\allowbreak Vector} classes. If \texttt{MatrixFree} has been configured by setting
\texttt{MatrixFree::AdditionalData::communicator\_sm} appropriately, then
\texttt{MatrixFree::create\_dof\_vector()} creates vectors that share
information among all processes on one node. As a consequence, the
\texttt{FEEvaluation} classes can access vector elements owned by
other processes and in certain cases local
communication can be skipped. To prevent race conditions, \texttt{MatrixFree} uses local
barriers at the beginning and the end of loops (\texttt{loop()}, \texttt{cell\_loop()}, \texttt{loop\_cell\_centric()}).

The new \texttt{step-76} tutorial program illustrates this latter case
in the context of the solution of the Euler equations. \texttt{step-76} reaches a
speed-up of 27\% compared to the 
original version, \texttt{step-67}, by using the new feature.
For more details and use of the feature in the library \texttt{hyper.deal}, see \citep{munch2020hyperdeal}.

@@

In addition, there are a number of new tutorial programs:
\begin{itemize}
\item \texttt{step-19} is an introductory demonstration of \dealii{}'s
  particle functionality. It solves the coupled problem of
  charged particles and an electric field, using a cathode tube as an
  example.

@@

\item \texttt{step-71} focused on automatic and symbolic
  differentiation (AD and SD, in short) as a tool to make solvers for complex, nonlinear
  problems possible. To this end, \dealii{} can interface to a number
  of AD and SD libraries, including Trilinos' Sacado package
  \cite{...}, ...\todo{which others??}, and SymEngine \cite{symengine-web-page}. The tutorial
  illustrates how these techniques can be used to compute derivatives
  of first a rather simple function, and then of the much more complex
  energy functions of two magneto-elastic and magneto-visco-elastic
  material formulations in which just the scalar energy functional
  takes up the better part of a page, and even first derivatives can
  only be computed with heroic effort.

\item \texttt{step-72} illustrates the use of automatic
  differentiation to simplify the computation of derivatives in the
  context of nonlinear partial differential equations where one needs
  to compute the Jacobian from the residual operator for efficient
  Newton iterations. \texttt{step-72} builds on the minimal surface
  solver \texttt{step-15} and replaces the hand-construction of the
  Jacobian by either computing it as the derivative of the residual,
  or alternatively as the second derivative of an energy functional
  that then also yields the residual itself.

@@

\item \texttt{step-77} is a program that illustrate \texttt{dealii}'s
  interfaces to the SUNDIALS library \cite{sundials}, and specifically
  the KINSOL nonlinear solver. Like the \texttt{step-72} program
  mentioned above, it is a variation of the minimal surface solver
  \texttt{step-15}: Instead of implementing the nonlinear Newton and
  line search loop ourselves, \texttt{step-77} relies on KINSOL for
  decisions such as when to rebuild the Jacobian matrix, when to
  actually solve linear systems with it, and how to form updates that
  drive the residual to convergence. The program illustrates the
  substantial savings that can be obtained by not re-inventing the
  wheel but instead building on an existing and well-tuned software
  such as KINSOL.

  \dealii{}'s interfaces were also updated to the latest SUNDIALS release, 5.7.
